---
output:
  html_document: default
  pdf_document: default
---
Linguistic data: quantitative analysis and visualization (Linguistic theory group)
Professor: Ilya Schurov | Student: Manuel Padilla-Moyano
Tittle: On the encoding of causal roles in Basque

# 1. Introduction
Languages have different morphosyntactical means to express causal relations. From a syntactic point of view, there are polipredicative (Мы не гуляли, потому что шёл дождь) and nominal causal constructions (Мы не гуляли из-за дожди). This study focuses on nominal causal constructions in Basque. Depending on factors not yet elucidated, when expressing causal relations, Basque writers choose different case markers: causal, instrumental, genitive, benefactive, or destinative. Until present, no attempt has been undertaken to describe the encoding of causal relations in Basque. My intuition is that the choice of the case marker depends on the causal semantic role (Cause, Force, Motive). Of course, there may be other factors, such as dialectal and diachronic ones. Beyond an initial description and quantification of the encoding of causal roles in nominal constructions, I will formulate some basic questions.

# 2. Typological ground
On the relationship between semantic roles Force and Cause. In Indo European languages Force is often coded as Instrument AND the instrumental expresses both Cause and Instrument [Luraghi 2010: 47]. Given that Basque is an isolated in terms of genetical relationships (albeit in a long lasting, intensive situation of contact with I-E languages), does Basque behave as an I-E language to this respect? Basque has both causal and instrumental cases. As an “I-E behavior” I expect Force to be coded by the instrumental case. As an “non-I-E behavior” I expect Force not to be coded by the instrumental case.

On the relationship between the semantic roles Cause and Motive [for which the label Reason is quite spread]. Although from a cognitive point of view Cause and Motive are significantly different, at least in the Indo-European languages they are not formally distinct: apparently, the normal situation is one where there is only one grammaticalized semantic role [Luraghi 2010: 54]. Does Basque behave as an I-E language to this respect or it encodes differently Cause and Reason? Basque has a causal case. As an “I-E behavior” I expect both Cause and Motive to be coded similarly. As an “non-I-E behavior” I expect Cause and Motive to be coded differently.

Future research should consider other semantic roles that overlap with the expression of causal relations (e.g. Purpose and Beneficiary).

# 3. 	Description of the data
NB: This project is conceived as the basis for a more extensive research that I will eventually conduct during the next year. Thus, in the case my data are not representative enough to prove or discard a given hypothesis in terms of statistical significance, I will hopefully have the opportunity to solve the problem.

I have compiled a small corpus of texts of historical Basque --I have also transcribed some of them--. The central part of the corpus consists of seven translations of the Gospel of Mark into four dialects, from 1571 to 2004. Five more texts have been added, in order to represent a more dialects at more periods. At this moment, the time span of this corpus covers from 1545 to 2004, and its size is 141,000 tokens.

Data consists of 2,338 observations. Each observation corresponds to one occurrence of any of the case-markers involved in the expression of the causal roles. For each observation I have annotated the following parameters: 1) number of verse, page, etc. where the example is found; 2) English translation --only for the versions of the Gospel--; 3) Basque example; 4) marker --suffix, postposition--; 5) allomorph, if any; 6) label of the case marker; 7) semantic role; 8) noun class --I have included interrogatives and connectors just for convenience--; 9) morphsyntactic remarks, if any; 10) semantic remarks, if any; 11) tittle of the text; 12) author; 13) year; 14) century; 15) period; 16) place; and 17) dialect.

```{r}
library(dplyr)
library(gplots)
library(tidyverse)
library(graphics)
```

```{r}
# Import the data and summarize them:
data <- read.csv("https://raw.githubusercontent.com/Leizarraga/encoding-causal-roles/master/encoding_causal_roles.csv", header = T, sep=';', dec = ".")
summary(data)
```

# 4.0 Starting point
Initial question: Are there any differences in the encoding of causal roles (Cause, Force, Motive) in Basque?
Derived questions: If yes, are such differences observable in all dialects? Are they stable throughout the history of Basque?
We could undertake the converse way too, i.e. analyze the semantic roles expressed by the case markers (causal, instrumental, genitive, benefactive): Which semantic roles does the causal (instrumental, genitive, benefactive) case-marker encode?
Are these roles uniformly represented among the different dialects? Are they uniformly represented throughout the different periods?

# 4.1. QUESTION 1: The encoding of Cause, Force and Motive
Are there differences in the encoding Cause, Motive, and Force in Basque? Here the hypotheses:
H0: encoding (Cause) = encoding (Motive) = encoding (Force)
H1: there are differences between any two of the three encodings

In this section we need to compare two categorical variables: which causal role is expressed by means of which case-marker. For this purpose, the suitable statistical tool is a chi-squared test, for which we need to use a table of contingency bigger than 2x2: at least 3 causal roles by 4 case-markers.

```{r}
# For convenience I simplify the data frame, so that only columns that are relevant to this paper are displayed:
library(dplyr)
data_light <- select(data, marker, case, role, dialect, century, year)
# And remove some rows which do not have any marker:
df <- subset(data_light, marker != "NOT_APPLY")
## Specific subsets of data for each semantic role studied in this paper:
Cause <- subset(df, role == "Cause")
Motive <- subset(df, role == "Motive")
Force <- subset(df, role == "Force")
```

```{r}
# Toward the contingency table
library(dplyr)
library(tidyverse)
# NB: Even though this step can be done in a simpler way, I will do some intermediary steps that might be useful later.
# 1. Create a new table wich displays semantic roles as columns, cases as rows, and number of occurrences:
case_role_NUMBERS <- df %>% count(case, role)
case_role <- case_role_NUMBERS %>% spread(role, n, fill = 0)
# 2. Select the roles and cases in which we are interested:
case_role_x3 <- select(case_role, case, Cause, Motive, Force)
case_role_4x3 <- filter(case_role_x3, case == "CAUS" | case == "INS" | case == "GEN" | case == "BEN")
case_role_4x3
# NB: It is necessary to remove the first column name ("role") so that work can be done with the contingency table. Honestly, I cannot manage to do it properly, and given the small size of the table I opt for creating a new txt file.
ctable_case_role <- read.delim("https://raw.githubusercontent.com/Leizarraga/encoding-causal-roles/master/ctable_case_role.txt", row.names = 1)
ctable_case_role
```

```{r}
# Visualization of the contingency table.
library(gplots)
# Convert the data as a table
dt <- as.table(as.matrix(ctable_case_role, row.names = 1))
# The function balloonplot() draws a graphical matrix where each cell contains a dot whose size reflects the relative magnitude of the corresponding component.
balloonplot(t(dt), main ="Causal_roles", xlab ="", ylab="",
            label = TRUE, show.margins = FALSE)
# show.margins = TRUE : marginal numbers will be shown.
# We can visualize a contingency table as a mosaic plot too.
library(graphics)
mosaicplot(dt, shade = TRUE, las=1,
           main = "causal_roles")
# The argument shade is used to color the graph
# The argument las = 1 produces horizontal labels
```
Blue color in the mosaicplot indicates that the observed value is higher than the expected value if the data were random. Conversely, red color specifies that the observed value is lower than the expected value if the data were random. 

This mosaic plot points to relevant differences between the encoding of Cause, Motive ad Force. The most remarkable are: 
1) the use of the genitive in the expression of Cause (indeed it hardly occurs with the other roles); 
2) the prominent use of the causal case in the expression of Motive;  
3) the prominence of the instrumental in the expression of Force; and 
4) the use of the benefactive case to encode Motive.

We can affirm that some cases are to a certain extent specialized in the expression of one of these causal roles, namely the benefactive case is (closely related to the expression of Motive), and the genitive (almost restricted to the expression of Cause).

Let us check the statistical significance of the remarks above.
```{r}
chisq <- chisq.test(ctable_case_role)
chisq
# The warning is perhaps due to the fact that some of our data are not enough (e.g. figures corresponding to Force), and therefore the approximations of p-value may not be right. Ideally, I could be able to collect new data so that numbers be higher, but the task must be undertaken in the future.
# Alternativelly we can perform a chi-squared test simulating p values, which gives a very slow p-value:
chisq <- chisq.test(ctable_case_role, simulate.p.value = TRUE)
chisq
# We can still use Fisher's Exact test. Again, p-value is very low.
# As the workspace needed is large, in order not to waste the time of the reader, I desactivate the code in this chunk. The results obtained were:
# fisher.test(ctable_role_case, workspace = 2e8, hybrid = FALSE)
#	Fisher's Exact Test for Count Data
# data:  ctable_role_case
# p-value < 2.2e-16
# alternative hypothesis: two.sided
```

In section 4.1 we have tested our hypothesis by different means, and we can reject the null-hypothesis, since the p-values obtained are of great statistical significance. Let us recall, our alternative hypothesis is that Basque encodes differently the causal roles termed as Cause, Motive, and Force.

# 4.2 QUESTION 2. Are these differences of encoding observable in all Basque dialects?
Null-hypothesis: the encoding of Cause, Force, and Motive is similar in Basque dialects.
Alternative hypothesis: the encoding of Cause, Force, and Motive differs in any pair of dialects.
For the sake of clarity and feasibility, we shall compare the encoding of only one role in different dialects and then repeat the statistical exercise for the other two roles.

# 4.2.1. Differences between dialects: Cause.
```{r}
library(dplyr)
library(tidyverse)
# 1. Create a new table wich displays dialects as columns, cases as rows, and number of occurrences:
Cause_case_dial_NUMBERS <- Cause %>% count(case, dialect)
Cause_case_dial <- Cause_case_dial_NUMBERS %>% spread(dialect, n, fill = 0)
# 2. Select the dialects and cases in which we are interested (we discard modern unified standard):
Cause_case_dial_x6 <- select(Cause_case_dial, case, Biscayan, Guipuscoan, High_Navarrese, Low_Navarrese, Labourdin, Souletin)
Cause_case_dial_4x6 <- filter(Cause_case_dial_x6, case == "CAUS" | case == "INS" | case == "GEN" | case == "BEN")
Cause_case_dial_4x6
# NB: I have the contingency table for Cause in a separate file:
ctable_Cause <- read.delim("https://raw.githubusercontent.com/Leizarraga/encoding-causal-roles/master/ctable_Cause.txt", row.names = 1)
ctable_Cause
```

Visualization of the contingency table.
```{r}
library(gplots)
dt <- as.table(as.matrix(ctable_Cause, row.names = 1))
balloonplot(t(dt), main ="Encoding of Cause in Basque dialects", xlab ="", ylab="",
            label = TRUE, show.margins = FALSE)
library(graphics)
mosaicplot(dt, shade = TRUE, las=1,
           main = "Encoding of Cause in Basque dialects")
```
The two graphs point to relevant differences in the encoding of Cause between the Basque dialects. The most outstanding facts are:
1) the prominence of the genitive in High Navarrese dialect; 
2) the more balanced distribution of case-markers Labourdin and Low Navarrese; 
3) the clarity of Biscayan data.
4) the marginality of the benefactive case in the expression of Cause.

Let us  check the statistical significance of the differences observed:
```{r}
chisq <- chisq.test(ctable_Cause)
chisq
chisq <- chisq.test(ctable_Cause, simulate.p.value = TRUE)
chisq
fisher.test(ctable_Cause, simulate.p.value = TRUE, hybrid = FALSE)
```
Even though some data might not be enough (e.g. figures for the benefactive), by simulating p-values or increasing the workspace we can test our contingency table. This way, the probability to obtain the distribution observed if the data were random is very small (p-value = 0.0004998), and therefore we can reject the null-hypothesis on the basis of statistics. Thus, we can affirm that there are differences in the encoding of the role Cause among Basque dialects. 

#  4.2.2. Differences between dialects: Motive.
```{r}
library(dplyr)
library(tidyverse)
Motive_case_dial_NUMBERS <- Motive %>% count(case, dialect)
Motive_case_dial <- Motive_case_dial_NUMBERS %>% spread(dialect, n, fill = 0)
Motive_case_dial_x6 <- select(Motive_case_dial, case, Biscayan, Guipuscoan, High_Navarrese, Low_Navarrese, Labourdin, Souletin)
Motive_case_dial_4x6 <- filter(Motive_case_dial_x6, case == "CAUS" | case == "INS" | case == "GEN" | case == "BEN")
Motive_case_dial_4x6
# NB: I have the contingency table for Cause in a separate file:
ctable_Motive <- read.delim("https://raw.githubusercontent.com/Leizarraga/encoding-causal-roles/master/ctable_Motive.txt", row.names = 1)
ctable_Motive
```

Visualization of the contingency table.
```{r}
library(gplots)
dt <- as.table(as.matrix(ctable_Motive, row.names = 1))
balloonplot(t(dt), main ="Encoding of Motive in Basque dialects", xlab ="", ylab="",
            label = TRUE, show.margins = FALSE)
library(graphics)
mosaicplot(dt, shade = TRUE, las=1,
           main = "Encoding of Motive in Basque dialects")
```
As for the encoding of Motive, the graphs point to much less difference between dialects than in the encoding of Cause. Two remarks can be made:
1) The benefactive case is unexpectedly used in Labourdin and Souletin dialects (especially in Labourdin); and 
2) In the High Navarresehe dialect the genitive has (indeed had) a wide use in the expression of Motive.

Statistical verification the differences observed.
```{r}
chisq <- chisq.test(ctable_Motive, simulate.p.value = TRUE)
chisq
fisher.test(ctable_Motive, simulate.p.value = TRUE, hybrid = FALSE)
```

# 4.2.3. Differences between dialects: Force.
```{r}
library(dplyr)
library(tidyverse)
Force_case_dial_NUMBERS <- Force %>% count(case, dialect)
Force_case_dial <- Force_case_dial_NUMBERS %>% spread(dialect, n, fill = 0)
# NB: there is no observation of the role Force in the High Navarrese dialect.
Force_case_dial_x5 <- select(Force_case_dial, case, Biscayan, Guipuscoan, Low_Navarrese, Labourdin, Souletin)
# NB: This time I do not include the genitive and the benefactive, as they are not attested in the encoding of Force (see in 4.3 Encoding_Force).
Force_case_dial_2x5 <- filter(Force_case_dial_x5, case == "CAUS" | case == "INS")
Force_case_dial_2x5
# NB: Contingency table of Force in a separate file.
ctable_Force <- read.delim("https://raw.githubusercontent.com/Leizarraga/encoding-causal-roles/master/ctable_Force.txt", row.names = 1)
ctable_Force
```

Visualization of the contingency table.
```{r}
library(gplots)
dt <- as.table(as.matrix(ctable_Force, row.names = 1))
balloonplot(t(dt), main ="Encoding of Force in Basque dialects", xlab ="", ylab="",
            label = TRUE, show.margins = FALSE)
library(graphics)
mosaicplot(dt, shade = TRUE, las=1,
           main = "Encoding of Force in Basque dialects")
```
Both the scarcity of the numbers and their distribution do not allow us to argue for significant differences in the expression of Force among the three dialects concerned. To say something, only in the Guipuscoan dialect things might be different, but the number of observations of Force for this dialect are negligible. In any case, it becomes clear that the instrumental case is by and large the most usual in the expression of Force.

Statistical significance.
```{r}
chisq <- chisq.test(ctable_Force, simulate.p.value = TRUE)
chisq
fisher.test(ctable_Force, simulate.p.value = TRUE, hybrid = FALSE)
```
Even simulating p-values, the statistical tools employed give p-values considerably over 0.5, which means that at the usual confidance level of 95% the null hypothesis cannot be rejected. I.e., there are no significant differences in the encoding of Force between Basque dialects.

# 4.3 QUESTION 3. Diachronic approach: are the differences of encoding (Cause, Motive, Force) stable throughout the textual history of Basque?

## At the beginning I did not manage to perform properly the logistic regression with the function summary().
## After incorporing it, it became clear that there is some very significant dependencies in the encoding of the three roles (indeed some of them show the highest vaue of significance).
## Thus, the text needs some reformulation.

Null-hypothesis: the encoding the role Cause, Force, and Motive is similar throughout five centuries.
Alternative hypothesis: the encoding of Cause, Force, and Motive differs at different periods during the last five centuries.

To answer the question we have to deal with numerical data (years or centuries, which are the basis for diachronic comparison) and categorical data (the case used to express a given semantic role). There is more than one possible mathematical model: we can consider Analysis of variances (ANOVA), or we can binarily encode the categorical variable, so that we can resort to logistic regressions. In this section I will opt for logistic regression, as it seems to me a poweful means to establish dependencies between two variables.

NB: The initial data frame had a column for "period", which for many specialits in diachrony is the most desirable parameter to compare data, but here I opt for working with years, for obvious reasons of usability in statistics.

Encoding of each role (without diachronic considerations):
```{r}
Encoding_Cause <- group_by(Cause, case)
summarise(Encoding_Cause,
  num = n()
)
Encoding_Motive <- group_by(Motive, case)
summarise(Encoding_Motive,
  num = n()
)
Encoding_Force <- group_by(Force, case)
summarise(Encoding_Force,
  num = n()
)
```

# 4.3.1. Diachronic approach: encoding of Cause.
```{r}
library(tidyverse)
library(dplyr)
# Create a new table for the role Cause with columns for year and for each one of the four cases:
Cause_case_year <- select(Encoding_Cause, year, case)
Cause_case_year_2 <- mutate(Cause_case_year, 
       CAUS = case == "CAUS",
       INS = case == "INS",
       GEN = case == "GEN",
       BEN = case == "BEN"
      )
Cause_case_year_3 <- subset(Cause_case_year_2, select = -case)
# Remove missing data:
Cause_case_year_4 <- na.omit(Cause_case_year_3)
# Arrange the new data frame by year:
Cause_case_year_final <- arrange(Cause_case_year_4, year)
# Binarize the data:
Cause_bin <- Cause_case_year_final * 1
```

Perform three logistic regressions (CAUS, INS, GEN; occurrences for BEN are negligible.)
```{r}
library(tidyverse)
lregr_Cause_CAUS <- glm(CAUS ~ year, data = Cause_bin, family = binomial(link = "logit"))
lregr_Cause_INS <- glm(INS ~ year, data = Cause_bin, family = binomial(link = "logit"))
lregr_Cause_GEN <- glm(GEN ~ year, data = Cause_bin, family = binomial(link = "logit"))
summary(lregr_Cause_CAUS)
summary(lregr_Cause_INS)
summary(lregr_Cause_GEN)
```

Visualization
```{r}
ggplot(data = Cause_bin, aes(x = year, y = CAUS)) +
  geom_point(aes(color = as.factor(CAUS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Cause_CAUS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Cause)",
       y = "Probability of CAUS") +
  theme(legend.position = "none")
ggplot(data = Cause_bin, aes(x = year, y = INS)) +
  geom_point(aes(color = as.factor(INS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Cause_INS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Cause)",
       y = "Probability of INS") +
  theme(legend.position = "none")
ggplot(data = Cause_bin, aes(x = year, y = GEN)) +
  geom_point(aes(color = as.factor(INS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Cause_GEN,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Cause)",
       y = "Probability of GEN") +
  theme(legend.position = "none")
```
For the encoding of Cause, the logistic model does not show dependence between the use of case-markers and the year. Considered by case-marker, there is total independence between the causal case (CAUS) and year. The other two cases analyzed show complementary relations of dependency: 
1) for the instrumental case, the regression model points to an increasing in its use by year (coefficient 0.001463);
2) for the genitive case, the dependency is defined in terms of negative coefficient (-0.0041), which means that every year the probability of GEN for the expression of Cause decreases. 

Obviously, the values for these coefficiennts would be much higher if we considered centuries instead of years.

```{r}
# Here I tried to visualize the relationship between P(GEN) and year, but something went wrong with the code.
ggplot(data = Cause_bin, aes(x = GEN, y = year, color = GEN)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.1) +
  theme_bw() +
  theme(legend.position = "null")
```

# 4.3.2. Diachronic approach: encoding of Motive
```{r}
library(tidyverse)
library(dplyr)
Motive_case_year <- select(Encoding_Motive, year, case)
Motive_case_year_2 <- mutate(Motive_case_year, 
       CAUS = case == "CAUS",
       INS = case == "INS",
       GEN = case == "GEN",
       BEN = case == "BEN"
      )
Motive_case_year_3 <- subset(Motive_case_year_2, select = -case)
Motive_case_year_4 <- na.omit(Motive_case_year_3)
Motive_case_year_final <- arrange(Motive_case_year_4, year)
Motive_bin <- Motive_case_year_final * 1
```

Perform four logistic regressions (CAUS, INS, GEN, BEN)
```{r}
library(tidyverse)
lregr_Motive_CAUS <- glm(CAUS ~ year, data = Motive_bin, family = binomial(link = "logit"))
lregr_Motive_INS <- glm(INS ~ year, data = Motive_bin, family = binomial(link = "logit"))
lregr_Motive_GEN <- glm(GEN ~ year, data = Motive_bin, family = binomial(link = "logit"))
lregr_Motive_BEN <- glm(BEN ~ year, data = Motive_bin, family = binomial(link = "logit"))
summary(lregr_Motive_CAUS)
summary(lregr_Motive_INS)
summary(lregr_Motive_GEN)
summary(lregr_Motive_BEN)
```

Visualization
```{r}
ggplot(data = Motive_bin, aes(x = year, y = CAUS)) +
  geom_point(aes(color = as.factor(CAUS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Motive_CAUS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Motive)",
       y = "Probability of CAUS") +
  theme(legend.position = "none")
ggplot(data = Motive_bin, aes(x = year, y = INS)) +
  geom_point(aes(color = as.factor(INS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Motive_INS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Motive)",
       y = "Probability of INS") +
  theme(legend.position = "none")
ggplot(data = Motive_bin, aes(x = year, y = GEN)) +
  geom_point(aes(color = as.factor(GEN)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Motive_GEN,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Motive)",
       y = "Probability of GEN") +
  theme(legend.position = "none")
ggplot(data = Motive_bin, aes(x = year, y = BEN)) +
  geom_point(aes(color = as.factor(BEN)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Motive_BEN,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Motive)",
       y = "Probability of BEN") +
  theme(legend.position = "none")
```
As for the encoding of Motive, the regression model clarifies some points: 
1) The dependency between P(CAUS) and year is clear, as the curve in the plot shows. 
2) The opposite is the tendency of P(BEN), which is increased by a coeeficient of 0.006408 per year.

# 4.3.3. Diachronic approach: encoding of Force
```{r}
library(tidyverse)
library(dplyr)
Force_case_year <- select(Encoding_Force, year, case)
Force_case_year_2 <- mutate(Force_case_year, 
       CAUS = case == "CAUS",
       INS = case == "INS",
       GEN = case == "GEN",
       BEN = case == "BEN"
      )
Force_case_year_3 <- subset(Force_case_year_2, select = -case)
Force_case_year_4 <- na.omit(Force_case_year_3)
Force_case_year_final <- arrange(Force_case_year_4, year)
Force_bin <- Force_case_year_final * 1
```

Perform two logistic regressions (CAUS, INS; there are no occurrences for GEN and BEN)
```{r}
library(tidyverse)
lregr_Force_CAUS <- glm(CAUS ~ year, data = Force_bin, family = binomial(link = "logit"))
lregr_Force_INS <- glm(INS ~ year, data = Force_bin, family = binomial(link = "logit"))
summary(lregr_Force_CAUS)
summary(lregr_Force_INS)
```

Visualization
```{r}
ggplot(data = Force_bin, aes(x = year, y = CAUS)) +
  geom_point(aes(color = as.factor(CAUS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Force_CAUS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Force)",
       y = "Probability of CAUS") +
  theme(legend.position = "none")
ggplot(data = Force_bin, aes(x = year, y = INS)) +
  geom_point(aes(color = as.factor(INS)), shape = 1) + 
  stat_function(fun = function(x){predict(lregr_Force_INS,
                                          newdata = data.frame(year = x),
                                          type = "response")}) +
  theme_bw() +
  labs(title = "Logistic regression (Force)",
       y = "Probability of INS") +
  theme(legend.position = "none")
```
Finally, and not surprisingly, the number of observations for the role Force is too small to obtain significant results. The picture is simple: this role is encoded by two case-markers, and the two show complementary evolution: while the use of the causal case gets increased in historical records, the use of the instrumental decreases to a similar extent. 

# 5. CONCLUSIONS
Based on more than 2,000 observations in a historical corpus of texts from all dialects, in this paper I have shown that Basque encodes differently three main causal roles (Cause, Motive, Force). As far as I am concerned, until present there is no description of the encoding of causal roles in Basque, particularly from a diachronic point of view. 

In the expression of Force Basque behaves as most Indo-European languages, since it encodes Force as Instrument (Basque has instrumental case). As for Cause and Motive, Basque presents certain divergences with respect to Indo-European languages: even if both roles are mostly encoded by means of the causal case, there are interesting tendencies. Our data show that in Northern dialects (Labourdin and Souletin) the benefactive case is increasingly used in the encoding of Motive. In terms of linguistic evolution, the encoding of Motive by the benefactive case is an innovation due to the contact with French --this point is out of the scope of this paper--.

Furthermore, the study revealed that the differences in the encoding of Cause, Motive and Force are observable also between Basque dialects. Finally, diachronical comparison points to some evolution in the expression of Cause, Motive and Force. For Cause, the use of the causal case remains stable during the historical period, the frequency of the instrumental goes up, and the genitive decays. For Motive, the use of the causal case shows a recessive tendency, and the benefactive emerges as an innovation. The expression of Force by the instrumental case remains more or less stable. 

# References
Luraghi, Silvia, 2010, “Adverbial Phrases” in Philip Baldi & Pierluigi Cuzzolin (eds.), New Perspectives on Historical Latin Syntax. Volume 2, De Gruyter Mouton, Berlin/New York.